---
title: "An Analysis of Education and Demographic Information on Immigrant Naturalization in America"
author: "Chaira Harder, Isabelle Elder, Sirohi Kumar"
format: pdf
editor: visual
---

## Abstract

The purpose of our analysis is to investigate the variables that determine if individuals choose to pursue US citizenship status after immigration. Understanding the variables that influence naturalization can inform future policy formation, aid in determining resource allocation, ensure legal and human rights considerations, and uphold equitable citizenship access. We used data from the IPUMS Higher Education database, which collects data from the Science and Engineers Statistical Data System, to analyze the determinants that increase likelihood of an individual gaining US citizenship status, including race/ethnicity, gender, employment, education level, and number of children. After filtering our data and using variable selection techniques to find an appropriate model, we found that race/ethnicity, gender, employment, and education level have significant impacts on the odds of an individual being a naturalized citizen. The more significant changes in someone’s odds of being a naturalized citizen are increases that come from being unemployed, then male. This research adds to a growing understanding of how different demographic and socioeconomic factors influence the citizenship process, which is especially relevant in this political climate.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, include = FALSE)
```

```{r}
# IMPORT LIBRARIES

library(dplyr)
library(ggplot2)
library(tidyverse)
library(tidytuesdayR)
library(corrplot)
library(MASS)
library(car)
# library(equatiomatic)
library(performance)
library(see)
library(tidyr)
```

## INTRODUCTION

The purpose of our analysis is to investigate the variables that determine if individuals choose to pursue US citizenship status after immigration. Understanding the variables that influence naturalization can inform future policy formation, aid in determining resource allocation, ensure legal and human rights considerations, and uphold equitable citizenship access. We used data from the IPUMS Higher Education database, which collects data from the Science and Engineers Statistical Data System, to analyze the determinants that increase likelihood of an individual gaining US citizenship status, including race/ethnicity, gender, employment, education level, and number of children. After filtering our data and using variable selection techniques to find an appropriate model, we found that race/ethnicity, gender, employment, and education level have significant impacts on the odds of an individual being a naturalized citizen. The more significant changes in someone’s odds of being a naturalized citizen are increases that come from being unemployed, then male. This research adds to a growing understanding of how different demographic and socioeconomic factors influence the citizenship process, which is especially relevant in this political climate.

## METHODS

<insert methods intro section here>

#### Data

This study uses data from IPUMS-Higher Education. This data set compiles information from the Science and Engineers Statistical Data System, the gold standard for longitudinally studying the education and employment of the US science and engineering workforce. Data from three different surveys sponsored by the National Science Foundation (NSF), including the National Surveys of College Graduates (NSCG), Recent College Graduates, and Doctorate Recipients, has been integrated from 1993 to present day. Of note, the NSRCG was discontinued in 2010, as survey items were folded into the other 2 surveys. However, the data set comprehensively brings together microdata to maintain continuity across variables. These surveys are administered biennially using a national widespread collection approach, where non-institutionalized individuals under the age of 76 who hold a bachelor’s degree or higher are asked but not required to respond. A single observation is defined as an eligible individual graduate in the US surveyed by one of the three NSF surveys, who is a naturalized US citizen.

```{r}
# NOTE: To load data, you must download both the extract's data and the DDI
# and also set the working directory to the folder with these files (or change the path below).

if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")

ddi <- read_ipums_ddi("highered_00002.xml")
data <- read_ipums_micro(ddi)
```

#### Variables

The outcome variable in our analysis is a binary variable indicating whether or not the respondent is a US citizen. This variable accounted for citizens by birth, leaving us with naturalized or non-citizen responses. We filtered the data to remove observations which had an NA value for citizenship status, which brought our data from 115,152 observations (total people) to 103,695 people (citizens). Gender is a binary categorical variable for male or female. Other variables of interest include race/ethnicity, a categorical variable that includes specific categories for Asian and White, with other race categories with fewer observations grouped into “Under-represented minorities” or “other”. Our educational variable is a categorical response to their highest degree received, as in a Bachelors, Masters, Doctorate, or Professional degree. Employment is grouped by employed, not employed, or not in the labor force.

```{r}
# WRANGLING
# remove identifier variables, as well as major field and citizen US binary
data <- data |>
  dplyr::select(-PERSONID, -YEAR, -SAMPLE, -SURID, -BTHUS, -NDGMEMG, -CTZUSIN)
```

```{r}
# WRANGLING
# reconstruct the CTZUS variable to be binary:
# for CTZUS == 3 (naturalized), we will make 1. All other categories (born abroad to US parents, US citizen from birth, Non-resident) will be 0
data$CTZUS <- ifelse(data$CTZUS == 3, 1, ifelse(data$CTZUS == 98, NA, 0))

# making '98' values --> NA in number of children variable CHTOT
data$CHTOT[data$CHTOT == 98] <- NA

# dim(data)
```

```{r}
# WRANGLING
# remove NA values from CTZUS
data = data |>
  dplyr::filter(!is.na(CTZUS))

# dim(data)
```

### Model Selection

<input model selection texts>

```{r}
# MODEL -- DELETE THIS??
# creating our SIMPLE logistic model

simpleModel = glm(CTZUS ~ DGRDG + LFSTAT + CHTOT, data = data, family = binomial)
summary(simpleModel)

# checking multicollinearity in our simple model
vif(simpleModel)
```

$$
\text{{CTZUS}} = -1.68659 + 0.12190 \text{{DGRDG}} - 0.01057 \text{{LFSTAT}} + 0.02518 \text{{CHTOT}})
$$

```{r}
# MODEL -- DELETE THIS??
# creating our COMPLEX logistic model

complexModel = glm(CTZUS ~ (DGRDG + LFSTAT + CHTOT) * (RACETH + GENDER), data = data, family = binomial)
summary(complexModel)
vif(complexModel)
```

$$
\text{{CTZUS}} = 1.68648 - 0.10993 \text{{DGRDG}} - 0.03052 \text{{LFSTAT}} -0.04997 \text{{CHTOT}} -1.57997 \text{{RACETH}} \\ -0.17444 \text{{GENDER}} + 0.05233 (\text{{DGRDG}} \times \text{{RACETH}}) - 0.05959 (\text{{LFSTAT}} \times \text{{RACETH}}) + \\ 0.01390 (\text{{CHTOT}} \times \text{{RACETH}}) + 0.05523 (\text{{DGRDG}} \times \text{{GENDER}}) + \\ 0.10552 (\text{{LFSTAT}} \times \text{{GENDER}}) + 0.03051 (\text{{CHTOT}} \times \text{{GENDER}})
$$

```{r}
# WRANGLING
# convert all variables except the outcome CTZUS: int to character
data[] <- lapply(data, as.character)
data$CTZUS <- as.numeric(data$CTZUS)
```

```{r}
# VARIABLE AND MODEL SELECTION
# creating a separate dataframe with no NA CHTOT values for stepwise AIC functionality

variableSelectionData = data |>
  filter(!is.na(CHTOT))
```

```{r}
# MODEL SELECTION
# creating the empty base model

empty_model <- glm(
  CTZUS ~ 1,
  data = variableSelectionData,
  family = binomial
)

summary(empty_model)
```

```{r}
# MODEL SELECTION
# creating the full base model

full_model <- glm(
  CTZUS ~ GENDER + RACETH + CHTOT + DGRDG + LFSTAT,
  data = variableSelectionData,
  family=binomial
)

summary(full_model)
```

```{r}
# MODEL SELECTION
# doing FORWARD StepAIC for prime model selection

stepAIC(
  object = empty_model,
  scope = list(lower = empty_model, upper = full_model),
  direction = "forward",
  test = "F",
  k = 2*qchisq(.95, df = 1)
)
```

```{r}
# MODEL SELECTION
# doing BACKWARD StepAIC for prime model selection

stepAIC(
  full_model,
  scope = list(lower = empty_model, upper = full_model),
  direction = "backward",
  test = "F",
  k = 2*qchisq(.95, df = 1)
)
```

```{r}
# MODEL SELECTION
# doing both StepAIC for prime model selection -- we can delete this, just did it to confirm.

stepAIC(
  object = empty_model,
  scope = list(lower = empty_model, upper = full_model),
  direction = "both",
  test = "F",
  k = 2*qchisq(.95, df = 1)
)
```

## RESULTS

```{r}
# MODEL SELECTION
# creating our best model based off of stepwise results

bestModel = glm(formula = CTZUS ~ RACETH + GENDER + DGRDG + LFSTAT, family = binomial, data = variableSelectionData)

# checking coefficient significance, AIC, and multicollinearity for best model
summary(bestModel)
vif(bestModel)

# dim(data)
```

$$
\log\left(\frac{p}{1 - p}\right) = 0.76524 - 3.71494 \cdot \text{RACETH2} - 2.38423 \cdot \text{RACETH3} + 0.33245 \cdot \text{GENDER2} + \\ 0.24602 \cdot \text{DGRDG2} + 0.30555 \cdot \text{DGRDG3} - 0.10952 \cdot \text{DGRDG4} + 0.48005 \cdot \text{LFSTAT2} - 0.08273 \cdot \text{LFSTAT3}
$$ Having generated this model, we can see that several of these coefficients are powerful predictors. Specifically, the race of the individual, whether they are male, whether they are unemployed, and if they have a doctorate or a professional degree significantly change the log-odds of them being a naturalized citizen. Further, the most significant changes in someone’s odds of being a naturalized citizen comes from whether someone is unemployed, followed by whether they’re male.

If we interpret these coefficients on the odds scale, we find that:

-   `Intercept`: The odds of being a naturalized citizen are $e^{0.76524} = 2.14951$ when you are an Asian female with a bachelor’s degree and you are employed.
-   `RACETH2`: the odds of being a naturalized citizen get $e^{-3.7194} = 0.02425$ times lower when you are white.
-   `RACETH3`: the odds of being a naturalized citizen get $e^{-2.38423} = 0.09218$ times lower when you are an underrepresented minority.
-   `GENDER2`: the odds of being a naturalized citizen increase by $e^{0.33245} = 1.39438$ times when you are a man.
-   `DGRDG2`: the odds of being a naturalized citizen increase by $e^{0.24602} = 1.27893$ when you have a Master’s degree.
-   `DGRDG3`: the odds of being a naturalized citizen increase by $e^{0.30555} = 1.35737$ when you have a Doctorate degree.
-   `DGRDG4`: The odds of being a naturalized citizen decrease by $e^{0.10952} =  1.11574$ when your education level is “Professional”. However, this is one of the least significant effects.
-   `LFSTAT2`: - The odds of being a naturalized citizen increase by $e^{0.48005} = 1.61616$ when you are unemployed.
-   `LFSTAT3`: The odds of being a naturalized citizen decrease by $e^{0.08273} = 1.08625$ when you are not in the labor force. This effect is also not significant.

Based on this data and this model, we can conclude that an Asian male, who has a Doctorate degree but is unemployed, has the highest odds of being a naturalized citizen. A female in the underrepresented minority group who has a Professional degree and is not in the labor force has the lowest odds of being a naturalized citizen, but this conclusion should be interpreted with caution given that the significance of our Professional and not being in the labor force coefficients are not significant.

## DISCUSSION

Initially, we predicted that all our predictor variables would have an impact. However, our variable selection method showed that the number of children did not have a significant effect in terms of predicting the odds of US citizenship. We also predicted that higher education levels would have a significant effect. Our results confirmed this, as the odds of being a naturalized citizen increased with a Masters and increased more with a Doctorate degree compared to a Bachelor's degree. These results seem likely given the prioritization of higher education, and how this could contribute to the citizenship process. Our results also confirmed that being male increased odds. We predicted that being an underrepresented minority would increase the likelihood of naturalization. However, our results showed that this actually significantly lowers the odds of being a naturalized citizen compared to being Asian, and being white also significantly lowers the odds. While these results were not exactly in line with our predictions, given the potential bias inherent to the naturalization process, these are not entirely shocking. Finally, we predicted being employed would increase the odds of being a naturalized citizen, but we found that unemployment increased the odds compared to being employed, and not being in the labor force decreased the odds compared to being employed. These results were the most surprising to us, as we expected employment to increase odds. We hypothesize that this could have some interaction with education, as the citizenship process could be looking to those with higher education to be seeking jobs and meet the criteria to fill them.

### LIMITATIONS

When choosing our model, the variable selection process necessitated removing observations with missing data, which decreased our number of datapoints due to how many observations were missing the total children variable. Due to the at-will responses to the surveys used to collect data used in this process, the sample may not necessarily be representative of the true educational and demographic characteristics of immigrants in the US, as certain factors may predispose people to answering the surveys. However, we think this analysis using this data likely still provides insight into potential impacts of relevant variables. Additionally, with the way race/ethnicity was grouped, the generalizations within these variables may mean details about these impacts are not fully captured within this data. All the variables in our dataset are categorical, with some redundancy, so some information may be vague. Our total children variable can only have 2 of 4 values used in the entire dataset because of redundancies in the question design, as it is not possible to have 02: 1-3 children and also 03: 2 or more children. Some of the categories, such as for DRDGR4 being “Professional”, are unclear in their meaning. We assume Professional means someone who is not necessarily holding a higher education degree, but has a technical degree for a specific profession, but this was unclear in the survey design and answers. Finally, our analysis does not prove causality between these variables, but rather, identifies potential correlations.

### FUTURE DIRECTION

Future research should consider further analysis into the employment aspect of the citizenship process, as this result was the most surprising. Additionally, expanding the race/ethnicity variable to include more groups could provide more information on how this variable truly impacts naturalization. Continued analysis of the factors that influence the naturalization process will add to the body of research, and can provide insight into the equitability of naturalization for different individuals.

\newpage

### References

1.  “World Population Prospects - Population Division.” United Nations, United Nations, 2022, population.un.org/wpp/.
2.  “U.S. Naturalization Policy.” CRS Reports, Congressional Research Service, 15 Apr. 2024, crsreports.congress.gov/product/pdf/R/R43366.
3.  Mossaad, Nadwa, et al. Determinants of Refugee Naturalization in the United States, Proceedings of the National Academy of Sciences of the United States of 4. America, 27 Aug. 2018, www.pnas.org/doi/abs/10.1073/pnas.1802711115.
4.  Batalova, Jeanne. “Frequently Requested Statistics on Immigrants and Immigration in the United States.” Migrationpolicy.Org, 4 Apr. 2024, www.migrationpolicy.org/article/frequently-requested-statistics-immigrants-and-immigration-united-states-2024#characteristics.
5.  U.S. and World Population Clock. United States Census Bureau. (2024, May 9). https://www.census.gov/popclock/

### IPUMS Citation

. Minnesota Population Center. IPUMS Higher Ed: Version 1.0 \[ImmigrationData\]. Minneapolis, MN: University of Minnesota, 2016. https://doi.org/10.18128/D100.V1.0

\newpage

## Data Appendix

#### Assumptions

**Linearity** Linearity is given because the predictors of our logistic model are all categorical.

**Independence** Given that our data samples a small portion of the population (about 115,000 out of 331 million Americans citizens), and that the data is surveying given characteristics (one person’s response will not affect another’s), we can assume that the data collected is independent.

**Randomness** This data comes from a database from various sources, but makes up a small portion of the total population. Further, the surveys collected data from individuals across the population (various colleges, etc). \<thus randomness is met.....\>
